Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-06-04 WholePlotTrajectories.Rmd
pdf

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric Marcon
    authsuperscript: 1
  - name: Bruno Hérault
    authsuperscript: 2 3
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: Cirad, Univ montpellier, UR Forests & Societies.
    address: >
      Montpellier,
      France.
  - affsuperscript: 3
    dptuniv: INPHB, Institut National Polytechnique Félix Houphouet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >   
    Understand the maintenance of forests biodiversity structure, functioning and dynamics is urgent to anticipate their fate in the global changing context.
    Communities diversity structure is assumed to rely on a constant regime of disturbance and peak at intermediate intensity.
    For tropical forests this intermediate disturbance hypothesis (IDH) remain debated as well as the completeness of forests resilience in all taxonomic and functoinal facets.

>>>>

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric Marcon
    authsuperscript: 1
  - name: Bruno Hérault
    authsuperscript: 2 3
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: Cirad, Univ montpellier, UR Forests & Societies.
    address: >
      Montpellier,
      France.
  - affsuperscript: 3
    dptuniv: INPHB, Institut National Polytechnique Félix Houphouet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >   
    Understand the maintenance of forests biodiversity structure, functioning and dynamics is urgent to anticipate their fate in the global changing context.
    Communities diversity structure is assumed to rely on a constant regime of disturbance and peak at intermediate intensity.
    For tropical forests this intermediate disturbance hypothesis (IDH) remain debated as well as the completeness of forests resilience in all taxonomic and functional facets.

### 2018-06-04 WholePlotTrajectories.tex
Graphs
pdf

\Keywords{mot-clés, séparés par des virgules} % Keywords - if you don't want any simply remove all the text between the curly brackets
\newcommand{\keywordname}{Keywords} % Defines the keywords heading name

@@

\Abstract{
Résumé de l'article.
}

@@

The large areas covered with tropical forests worldwide hold crucial
economic, social and cultural values. They provide wood and multiple
non-timber forest products, shelter a diversified fauna, regulate the
local climate, support the carbon, water and nutrient cycles, and ensure
cultural and human well-being. The simultaneous increase of forests
products demand and the substantial climatic changes currently heighten
the pressure on remaining forests
\citep{Gibson2011a, Morales-Hidalgo2015}, threatening the maintenance of
communities structure, composition and functioning and their dynamics in
space and time \citep{Anderson-Teixeira2013, Sist2015}.

In tropical forest, ecological communities are constantly re-shaped by
the natural disturbance events changing both abiotic environment,
through the fluxes of light, heat and water, and biotic interactions and
competitive pressure \citep{Goulamoussene2017}. The cornerstone of
tropical forests ecology is to understand the mechanisms and the
determinants of ecosystems response to disturbance
\citep{White2001, Chazdon2003a}. For now, this has been largely studied
through structural parameters, rapid and convenient to measure, as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. These structural parameters were
thereafter consistently modeled and gave important insights into the
maintenance of ecosystems processes and services
\citep{Denslow2000, Blanc2009}. However the response of forests
diversity in tree species remains unclear, albeit it determines the
productivity, stability and functioning of ecosystems
\citep[\citet{Liang2016}]{Tilman2014} and would be most probably
impacted by the changes following disturbance \citep{Baraloto2012a}.

In the short-term disturbance demonstrated negligible or even positive
impacts on communities diversity, which have been formalized by the
intermediate disturbance hypothesis (IDH) stating a maximized species
diversity at intermediate disturbance intensity
\citep{Molino2001, Kariuki2006a, Berry2008a}. Still, validations of the
IDH remain scarce in the long term and mainly rely on species richness
analyses that gives limited or misleading information on forests
recovery and functioning \citep{Martin2015, Chaudhary2016}. More
releveant monitoring would encompass communities composition, crucial
for conservation issues, and complete diversity profiles encompassing
species evenness in addition to their richness, that reveals ecological
rules shaping communities
\citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, account
for species biological attributes and role in the ecosystem is essential
to understand the correlations between ecosystems biodiversity,
functioning and environmental constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. In that
respect functional approaches based on major traits related to species
ecology and performance was largely adopted
\citep{Diaz2005, Villeger2008a}, for example highlighting in tropical
rainforests the environmental filters fostering fast growing species
with efficient resources acquisition after disturbance
\citep{Molino2001, Haddad2008}. Post-disturbance trajectories then
corresponded to a shift from ``conservative'' slow-growing species
dealing with scarce resources dominating before disturbance, to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{TerSteege2001, Reich2014, Herault2011}. This
shift translated into consistent trajectories for key functional traits
related to resource acquisition (leaf area, density and chlorophyll
content, and stem specific gravity and bark thickness), tree growth and
reproduction and life history traits (seed mass and maximum height)
\citep{Wright2004, Westoby2006a, Chave2009b}.

Functional and taxonomic trajectories are both essentiall to fully
assess communities response to disturbance. Both relate to distinct
ecological rules and processes and might then be decoupled as it was
observed already in tropical forests
\citep{Lohbeck2015, Guariguata2001}. Although taxonomic and functional
diversity are complementary, they are combine into communities
functional redundancy that measures the amount of species sharing same
trait values \citep{Carmona2016}. Functional redundancy then explicits
the link between otherwise decoupled taxonomic and functional
diversities and is central for communities description. Besides,
functional redundancy determines communities resilience as high
redundancy, like in highly diverse tropical forests
\citep{Bellwood2006}, mitigates the impacts of species removal on
ecosystem functioning \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Grasp all facets of communities response to disturbance amount to
disentangle the taxonomic and functional trajectories in diversity and
composition. These trajectories would highlight the ecological rules
constraining, or not, communities dynamics towards a recovery of initial
composition, diversity and functioning. They would therefore clarify the
tenants in the long term of the debated Intermediate Disturbance
Hypothesis for tropical forests, which would be as much insights for
future adaptive conservation strategies \citep{Adler2007}. Here we
monitored over 30 years the response of 75 ha of forests plots set up on
a gradient of disturbance intensity, from 10 to 60\% of ecosystem
biomass removed. We made use of a large functional traits database
browsing major leaf, stem and seed traits and species maximum height to
draw the trajectories over time of communities taxonomic and functional
composition and diversity. Specifically, we (i) questioned the coupling
between taxonomic and functional response to disturbance and identified
the underlying assembly processes, which allowed to (ii) clarify the
validity of the IDH in the long term for tropical forest and (iii)
question the completeness of communities recovery regarding their
functional redundancy.

@@

The experiment is a network of twelve 6.25ha plots that underwent a
gradient of three logging, thinning and fuelwood cutting treatments
(Table \ref{tab:Tab1}). Disturbance treatments were attributed according
to a randomized plot design with three replicate blocks of four plots.
The disturbance corresponds to averages of 10 trees removed per hectare
with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1),
32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees above 40
cm DBH for treatment 3 (T3). Treatments T2 and T3 besides included the
thinning of trees by poison girdling \citep{Blanc2009}. The disturbance
intensity was measured as the percentage of aboveground biomass (\%AGB)
lost between the first inventory in 1984 and five years after
disturbance (ref to be found) measured with the BIOMASS R package
\citep{Biomass2018}.

@@

The taxonomic diversity was assessed through Richness and the Hill
number translation of Shannon and Simpson indices \citep{Hill1973}.
These three indices belong to the set of HCDT or generalized entropy,
respectively corresponding to the 0, 1 and 2 order of diversity (q),
which proved well suited for diversity studies
\citep{Patil1982, Tothmeresz1995}. The functional diversity was reported
using the Rao index of quadratic entropy which combines species
abundance distribution and average pairwise dissimilarity based on all
functional and life traits.

@@

The impact of disturbance was examined specifically through the linear
correlation between the intial \%AGB removed and the Simpson and Rao
diversities (diversities of order 2) after 10, 20 and 30 years
\ref{fig:IDHplot}. The correlation with disturbance intensity was weak
for the Simpson diversity (\(R^2<0.25\)) and only valid from 20 years
after disturbance but it was much stronger for the Rao diversity
(\(0.55<R^2<0.75\)) for all the time studied. Slope of linear
correlations, reflecting the impact of disturbance, was the highest 20
years after disturbance.

@@

Considering the functional redundancy restricted to the functional space
of the initial inventory, all disturbed plots followed similar
decreasing humped-shaped trajectories (@ref(fig:RedFun\_rest)). The
maximum redundancy loss was positively correlated with the disturbance
intensity (XX spearman to be measured) and the initial value had not
recovered for any disturbed communities.

@@

\subsection{Decoupled recovery of communities taxonomic and functional
characteristics}\label{decoupled-recovery-of-communities-taxonomic-and-functional-characteristics}

Both communities taxonomic and functional diversity and composition
proved resilient, following similar humped-back trajectories with a
return towards initial values. The resilience of communities functional
characteristics, the most direct link between biodiversity and ecosystem
functioning \citep{Diaz2005}, meant a consistent recovery of ecosystem
processes in the long term \citep{Guariguata2001}. The resilience of
communities taxonomic characteristics, in turn, meant the maintenance of
their initial differences in composition and structure. It suggested
that communities response to disturbance were somehow constraint to
converge towards determined compositions
\citep{Hubbell1999, Molino2001, Anderson2007, Baraloto2012a}.

Although both communities taxonomic and functional characteristics
proved resilient and followed similar humped-back trajectories, the
taxonomic recovery systematically lagged behind the corresponding
functional dynamics. Such delay between functional and taxonomic
dynamics has already been observed for grasslands
\citep{Tilman1997, Mouillot2011} and more recently for tropical forests
\citep{Lohbeck2015, Guariguata2001}. According to the ``vegetation
quantity effect'' \citep{Grime1998} functional trajectories rely on the
pool of dominant species, which diversity and evenness were enhanced
after disturbance, and which rapidly restored the functional diversity
and composition. However, communities evenness remained high so
infrequent species still missed to the taxonomic recovery that
mechanically lagged behind, all the more so that unrecovered species
would be functionally redundant and probably undergo competition
processes.

@@

Validating the IDH, communities trajectories confirmed the diversity
increase after disturbance through an enhanced growth of otherwise less
favored species. This was however completely clear regarding the
functional diversity but much more blurry for communities taxonomic
characteristics.

The taxonomic richness was weakly or negatively impacted by intense
disturbance, as observed on several post logging surveys
\citep{Cannon1998, Baraloto2012a}, while it substantially increased
after low intensity disturbance. Disturbance also enhanced taxonomic
evenness, but its intensity was only weakly and lately correlated to
communities eveness (\emph{i.e.} Simpson diversity). Contrastingly,
disturbance intensity consistently predicted the significant increase of
communities functional diversity for 30 years. Disturbance impacts
involve specific turnover within communities, either among
pre-disturbance survivors or among newly recruited trees. As the
composition of old-growth survivors proved to mirror initial communities
\citep{Herault2018}, disturbance would impact trees recruitment through
the enhanced growth and survival of previously infrequent species and
functional characteristics. Consistently, disturbance resulted in
increasing taxonomic dissimilarity compared to pre-disturbance
communities and functional shifts towards resource-acquisitive
strategies (sharp increase in the SLA, leaf thickness and bark thickness
and decrease in wood density, leaf toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. Disturbance then caused a
reorganization of the typical high dominance structure of hyperdiverse
mature forests after disturbance, benefiting to pioneers and light
demanding species. Likely, the changes in abiotic environment and
competitive pressure favored pioneers which outcompete others in non
limiting resources but are excluded in mature forests by long-lived,
more resistant and shade tolerant species. Therefore, consistently with
communities dynamics after disturbance relied on species functional
strategy and corresponding ability to fill the environmental niches made
available by disturbance. Recruited species then mixed with
pre-disturbance ones, from which they differed, and constituted a
community all the more diversified that the disturbance was intense
\citep{Molino2001}.

\subsection{The functional redundancu and the resilience of
communities}\label{the-functional-redundancu-and-the-resilience-of-communities}

Both the lag between taxonomic and functional recovery and the middling
consistency of the IDH regarding communities taxonomic diversity entail
a central role of functional redundancy. Functional redundancy,
determinant of communities resilience
\citep{Trenbath1999, Elmqvist2003, Diaz2005}, seemed not to have
recovered 30 years after disturbance.

Although disturbance would be mitigated by communities functional
redundancy, we expect the redundancy itself to be reduced or
re-organized within the functional space. Globally, disturbed
communities did not follow consistent trajectories after disturbance and
30 years after disturbance, irrespective of the intensity, disturbed
communities displayed lower, similar or higher functional redundancy.
The redundancy within the functional space of the initial community
however clearly followed humped-back trajectories which decrease was
determined by the disturbance intensity and which return to initial
state was not observed for any disturbed community. Not only did the
functional redundancy and hence the resilience of pre-disturbance
community decrease, but also did the ones of disturbance-specific
communities increase. It means higher chances to have long lasting or
self-maintained compositional changes towards disturbance resistant
species, lianas or epiphytes \citep{Haddad2008, Burslem2000, Martin2013}
and thus highly question forest's resilience \citep{Chazdon2003a}.
Specificaly, this would impair species contingent to undisturbed
forests, threatening their maintenance, and run the risk to loose
cornerstone species and trigger unexpected ecological consequences
\citep{Jones1994, Diaz2005, Gardner2007}.

@@

our study disentangled tropical forests functional and taxonomic
decoupled response to disturbance, with a rapid recovery of communities
functioning but slower and more variable taxonomic dynamics.
Consistently with the IDH functional trajectories were constrained by
environmental pressures favoring pioneers and light-demanding species
after disturbance, but taxonomic dynamics were more stochastic as they
involved infrequent speciesfunctoinally redundant with the dominants and
thus under competitive pressure. While communities functioning rapidly
recovered, taxonomic trajectories, although consistent and converging
towards full recovery, were much slower and entailed persisting
alteration of communities functional redundancy. the trajectories
therefore suggested a potential sustainability of tropical forests in
the face of quite intense disturbance but only when followed by long
recovery periods, longer than 30 years \citep{Gourlet-Fleury2005}. This
study besides highlighted the central role of recruitment processes
underlying the IDH mechanisms and holding the whole community response
to disturbance.

>>>>

\Keywords{Taxonomic and Functional Biodiversity, Neotropical Forests, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Long-term Resilience} % Keywords - if you don't want any simply remove all the text between the curly brackets
\newcommand{\keywordname}{Keywords} % Defines the keywords heading name

@@

\Abstract{
Understand the maintenance of forests biodiversity structure,
functioning and dynamics is urgent to anticipate their fate in the
global changing context. Communities diversity structure is assumed to
rely on a constant regime of disturbance and peak at intermediate
intensity. For tropical forests this intermediate disturbance hypothesis
(IDH) remain debated as well as the completeness of forests resilience
in all taxonomic and functional facets.
}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local climate, support the carbon, water and nutrient
cycles, and ensure cultural and human well-being. The simultaneous
increased demand in forests products and the substantial climatic
changes currently heighten the pressure on remaining forests
\citep{Gibson2011a, Morales-Hidalgo2015}, threatening the maintenance of
communities structure, composition and functioning and their dynamics in
space and time \citep{Anderson-Teixeira2013, Sist2015}.

In tropical forest, ecological communities are constantly re-shaped by
natural disturbance events that change both abiotic environment, through
the fluxes of light, heat and water \citep{Goulamoussene2017}, and
biotic interactions like competition among species \citep{Herault2018}.
The cornerstone of tropical forests ecology is to understand the
mechanisms and the determinants of ecosystems response to disturbance
\citep{White2001, Chazdon2003a}. For now, this has been largely studied
through structural parameters, rapid and convenient to measure, as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. These structural parameters were
thereafter sucessfully modeled, giving important insights into the
maintenance of ecosystems processes and services
\citep{Denslow2000, Blanc2009}. However the response of forests
diversity in tree species remains unclear, albeit it determines the
productivity, stability and functioning of ecosystems
\citep[\citet{Liang2016}]{Tilman2014} and would be most probably
impacted by the changes following disturbance \citep{Baraloto2012a}.

In the short-term disturbance demonstrated negligible or even positive
impacts on communities diversity, which have been formalized by the
intermediate disturbance hypothesis (IDH) stating a maximized species
diversity at intermediate disturbance intensity
\citep{Molino2001, Kariuki2006a, Berry2008a}. Still, validations of the
IDH remain scarce in the long term and mainly rely on species richness
analyses that give limited or misleading information on forests recovery
and functioning \citep{Martin2015, Chaudhary2016}. More releveant
monitoring would encompass communities composition, crucial for
conservation issues, and complete diversity profiles integrating both
species richness and evenness, to reveal the ecological rules
structuring communities \citep{Magurran1988, Lavorel2002, Bellwood2006}.
Furthermore, the functional approach that accounts for species
biological attributes and role in the ecosystem would be essential to
understand the correlations between ecosystems biodiversity, functioning
and environmental constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. The
functional trait-based approach focused on major traits related to
species ecology and performance was sucessfully adopted
\citep{Diaz2005, Villeger2008a}, for example in highlighting
deterministic processes in tropical rainforests that foster fast growing
species with efficient resources acquisition after disturbance
\citep{Molino2001, Haddad2008, Ruger2009}. Communities then shifted from
a dominance of ``conservative'' slow-growing species dealing with scarce
resources to ``acquisitive'' fast-growing species with rapid and
efficient use of abundant resources
\citep{TerSteege2001, Reich2014, Herault2011}. This shift translated
into consistent trajectories for key functional traits related to
resource acquisition (leaf area, density and chlorophyll content, and
stem specific gravity and bark thickness), tree growth and reproduction
and life history traits (seed mass and maximum height)
\citep{Wright2004, Westoby2006a, Chave2009b}.

A combination of taxonomic and functional approaches are essential to
fully assess communities response to disturbance and differences in
their respective trajectories are insightful of underlying mechanisms
\citep{Lohbeck2015, Guariguata2001}. Although taxonomic and functional
diversity are complementary and sometimes decouples, they are combined
in the measure of communities functional redundancy that quantifies the
amount of shared trait values among species \citep{Carmona2016}.
Functional redundancy is determinant of communities resilience as high
redundancy, like in highly diverse tropical forests
\citep{Bellwood2006}, mitigates the impacts of species removal on
ecosystem functioning \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Grasp all facets of communities response to disturbance comes to examine
the taxonomic and functional trajectories in diversity and composition.
These trajectories would highlight the ecological rules constraining or
not communities dynamics towards the recovery of initial composition,
diversity and functioning. They would therefore clarify the tenants in
the long term of the debated Intermediate Disturbance Hypothesis for
tropical forests, a cruicial point for future adaptive conservation
strategies \citep{Adler2007}. Here we monitored over 30 years the
response of 75 ha of forests plots set up on a gradient of disturbance
intensity, from 10 to 60\% of ecosystem biomass removed. We made use of
a large functional traits database browsing major leaf, stem and seed
functional traits and species maximum height to draw the trajectories
over time of communities taxonomic and functional composition and
diversity. Specifically, we (i) questioned the coupling between
taxonomic and functional response to disturbance and identified the
underlying assembly processes, which allowed to (ii) clarify the
validity of the IDH in the long term for tropical forest and (iii)
question the completeness of communities recovery regarding their
functional redundancy.

@@

The experiment is a network of twelve 6.25ha plots that underwent a
gradient of three logging, thinning and fuelwood cutting treatments
(Table \ref{tab:Tab1}). Disturbance treatments were attributed according
to a randomized plot design with three replicate blocks of four plots.
The disturbance corresponds to averages of 10 trees removed per hectare
with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1),
32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees/ha above
40 cm DBH for treatment 3 (T3). Treatments T2 and T3 besides included
the thinning of trees by poison girdling \citep{Schmitt1989, Blanc2009}.
The disturbance intensity was measured as the percentage of aboveground
biomass (\%AGB) lost between the first inventory in 1984 and five years
after disturbance (ref to be found) measured with the BIOMASS R package
\citep{Biomass2018}.

@@

The taxonomic diversity was assessed through Richness and the Hill
number translation of Shannon and Simpson indices \citep{Hill1973}.
These three indices belong to the set of HCDT or generalized entropy,
respectively corresponding to the 0, 1 and 2 order of diversity (q),
recomended for diversity studies \citep{Patil1982, Tothmeresz1995}. The
functional diversity was reported using the Rao index of quadratic
entropy which combines species abundance distribution and average
pairwise dissimilarity based on all functional and life traits.

@@

The impact of disturbance was examined specifically through the linear
correlation between the intial \%AGB removed and the Simpson and Rao
diversities (diversities of order 2) after 10, 20 and 30 years
\ref{fig:IDHplot}. The correlation with disturbance intensity was weak
for the Simpson diversity (\(R^2<0.25\)) and only valid from 20 years
after disturbance but it was much stronger for the Rao diversity
(\(0.60<R^2<0.75\)) for all the time studied. Slope of linear
correlations, reflecting the impact of disturbance, was the highest 20
years after disturbance.

@@

Considering the functional redundancy restricted to the functional space
of the initial inventory, all disturbed plots followed similar
decreasing humped shaped trajectories (@ref(fig:RedFun\_rest)). The
maximum redundancy loss was positively correlated with the disturbance
intensity (XX spearman to be measured) and the initial value had not
recovered for any disturbed communities.

@@

\subsection{Decoupled taxonomic and functional
trajectories}\label{decoupled-taxonomic-and-functional-trajectories}

Both communities taxonomic and functional diversity and composition
proved resilient, following similar humped shaped trajectories returning
towards initial values. The resilience of communities functional
structure, that is the most direct link between biodiversity and
ecosystem functioning \citep{Diaz2005}, meant a consistent recovery of
ecosystem processes in the long term \citep{Guariguata2001}. The
resilience of communities taxonomy meant the maintenance of their
initial differences in composition and diversity, suggesting the
existence of multiple stable equilibria as assumed for highly diverse
and productive ecosystems \citep{Chase2003}. It besides suggested that
recovery trajectories depend on the initial composition eventually
restored that would the the asymptot of the humped shaped dynamics
\citep{Hubbell1999, Molino2001, Anderson2007, Baraloto2012a}.

Although both communities taxonomic and functional characteristics
proved resilient and followed similar humped shaped trajectories, the
taxonomic recovery systematically lagged behind the functional one. The
decoupling between functional and taxonomic dynamics has already been
observed for grasslands \citep{Tilman1997, Mouillot2011} and more
recently for tropical forests \citep{Lohbeck2015, Guariguata2001}.
According to the ``vegetation quantity effect'' \citep{Grime1998}
functional trajectories rely on the pool of dominant species, which
diversity and evenness were enhanced after disturbance, and which
rapidly recovered their initial functional structure. However,
communities evenness remained high so infrequent species still missed to
the taxonomic recovery that proved all the more longer to recover that
unrecovered species would be functionally redundant and probably
underwent competition processes.

@@

Validating the IDH, communities trajectories confirmed the diversity
increase after disturbance through an enhanced growth of otherwise less
favored species. This was strong regarding the functional diversity but
more blurry regarding their taxonomic structure.

The taxonomic richness was weakly or negatively impacted by intense
disturbance, as observed on several post logging surveys
\citep{Cannon1998, Baraloto2012a}, while it substantially increased
after low intensity disturbance. Disturbance also enhanced taxonomic
evenness, but the disturbance intensity was only weakly and lately
correlated to communities eveness (\emph{i.e.} Simpson diversity).
Contrastingly, disturbance intensity consistently predicted the
significant increase of communities functional diversity for 30 years.
Disturbance entailed specific turnover within communities, either among
pre-disturbance survivors or among newly recruited trees. As the
composition of old-growth survivors proved to mirror initial communities
\citep{Herault2018}, disturbance likely impacted trees recruitment
through the enhanced growth and survival of previously infrequent
species and functional types. Consistently, disturbance resulted in
increasing taxonomic dissimilarity compared to pre-disturbance
communities and significant functional shifts towards
resource-acquisitive strategies (sharp increase in the SLA, leaf
thickness and bark thickness and decrease in wood density, leaf
toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. Disturbance then caused a
reorganization of the typical high dominance structure of hyperdiverse
mature forests after disturbance, benefiting to pioneers and light
demanding species. Likely, the changes in abiotic environment and
competitive pressure favored pioneers which outcompete other species in
non limiting resources but are excluded in mature forests through the
cmpetition with long-lived, more resistant and shade bearers. Therefore,
consistently with the IDH the trajectories after disturbance relied on
the environmental niches made available by disturbance filled by species
from a restrictied functional range. Recruited species therefore
differed from pre-disturbance ones, constituting a community all the
more diversified that the disturbance was intense \citep{Molino2001}.

\subsection{The functional redundancy, key of communities
resilience}\label{the-functional-redundancy-key-of-communities-resilience}

Both the lag between taxonomic and functional recovery and the middling
consistency of the IDH regarding communities taxonomic diversity made
central the role of functional redundancy. Functional redundancy,
determinant of communities resilience
\citep{Trenbath1999, Elmqvist2003, Diaz2005}, seemed not to have
recovered 30 years after disturbance.

Communities with high functional redundancy are expected to be able to
lose species with a limited impact on their functional structure
\citep{Carmona2016}. Functional redundancy therefore mitigates the
impacts of disturbance, but we expect the redundancy itself to be
reduced or re-organized within the functional space. Although the
overall functional redundancy did not follow consistent trajectories
after disturbance, the redundancy restricted to the functional space of
the initial community clearly followed humped shaped trajectories. This
restricted redundancy decreased according to the disturbance intensity
before returning towards the initial value but had not recovered for any
disturbed community after 30 years. The recovery of the redundancy
within the initial functional space explained the hampered recovery of
infrequent species specific to old-growth forests that would be
functionally redundant and therefore underwent competitive pressures
limiting species functional similarity \citep{Mayfield2010}. The
unachieved humped shaped redundancy trajectory meant a decrease of the
resilience of pre-disturbance communities and an increase in the
resilience of disturbance-specific communities (corresponding to lower
richness and higher dominance of pioneer species). After disturbance the
chances were higher to have long lasting or self-maintained
compositional changes in favor of disturbance resistant species, lianas
or epiphytes \citep{Haddad2008, Burslem2000, Martin2013}, which highly
questioned the completeness of the resilience \citep{Chazdon2003a}.
Specificaly, species contingent to undisturbed forests might be
impaired, threatening their maintenance, the risk to loose cornerstone
species and trigger unexpected ecological consequences would be higher
\citep{Jones1994, Diaz2005, Gardner2007}.

@@

our study defined the decoupled functional and taxonomic trajectories of
tropical rainforests after disturbance. These trajectories shed light on
their rapid functional recovery but slower and more variable taxonomic
dynamics. Consistently with the IDH, functional trajectories were
constrained by determined processes favoring pioneers and
light-demanding species after disturbance. The taxonomic dynamics
however proved more stochastic due to the recovery of infrequent,
functionally redundant species hampered by limiting similarity
processes. Although communities functioning rapidly recovered, their
taxonomic structure and functional redundancy remained altered for
several decades. The trajectories therefore demonstrated a decades-long
resilience of tropical forests in the face of quite intense disturbance
\citep{Gourlet-Fleury2005}. Within the mechanisms of this resilience the
recruitment processes proved central, in accordance with the IDH, and
would deserved closer focus to clarify their determinisms and ecological
rules. The disturbance range however stayed within the spectrum of
selective logging, with a forest cover remaining all along the
experiment, and the response mechanisms would probably be much different
after harder disturbance.
